A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-05-29 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Added some missing references flagged during proof-reading for typesetting.

Whether results from laboratory experiments will hold when the virus circulates in the community remains uncertain. At a global scale, de Ángel Solá et al. [-@deangel2020weathering] see less risk from SARS-CoV2 in the Caribean Basin; on the other hand, Coelho et al. [@Coelho2020exponential] warn that at least during the exponential phase, expansion of the virus is not driven by climate. Similarly, whereas Araujo and Naimi [-@Araujo2020spread] argue that spread of SARS-CoV2 will likely be constrained by climate, Harbert et al. [-@Harbert2020spatial] remain unconvinced that spatial modelling can currently discriminate the distribution of the disease on the basis of climate, at least in the United States. Yao et al. [-@Yao2020association], examined data from China and came to the conclusion that neither temperature nor ultraviolet indices had an association with transmission of COVID-19. This is despite previous research that has linked less exposure to UVB radiation to higher prevalence and severity of acute respiratory tract infections (Zittermann et al. 2016; Dąbrowska-Leonik et al. 2018; Dinlen et al. 2016; Mathyssen et al. 2017; Esposito and Lelii 2015; Jat 2017; Moriyama, Hugentobler, and Iwasaki 2020).

@@

As is the case with cross-sectional data, it is possible to test the residuals of Model (\ref{eq:sur-sim-block}) for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial]. When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation]. In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor. Spatial analytical approaches were used to understand contagion-difussion processes in the case of infectious disease in general [e.g., @Cliff1998detecting] and the 2003-2004 SARS outbreak in particular [e.g., @Meng2005understanding; @Cao2010spatio]. While we are mindful of the same caveat that the novel SARS-CoV2 may not follow the patterns of previous diseases, there is still evidence from the United States that COVID-19 displays spatial patterns that are consistent with a diffusion process [@Desjardins2020rapid]. For this reason, the spatial lag model is appropriate to model incidence of COVID-19 geographically, since it accounts for potential spatial patterns that result from a process of contagion, as explained next.

>>>>

Whether results from laboratory experiments will hold when the virus circulates in the community remains uncertain. At a global scale, de Ángel Solá et al. [-@deangel2020weathering] see less risk from SARS-CoV2 in the Caribean Basin; on the other hand, Coelho et al. [@Coelho2020exponential] warn that at least during the exponential phase, expansion of the virus is not driven by climate. Similarly, whereas Araujo and Naimi [-@Araujo2020spread] argue that spread of SARS-CoV2 will likely be constrained by climate, Harbert et al. [-@Harbert2020spatial] remain unconvinced that spatial modelling can currently discriminate the distribution of the disease on the basis of climate, at least in the United States. Yao et al. [-@Yao2020association], examined data from China and came to the conclusion that neither temperature nor ultraviolet indices had an association with transmission of COVID-19. This is despite previous research that has linked less exposure to UVB radiation to higher prevalence and severity of acute respiratory tract infections [@Zittermann2016vitamin; @Dąbrowska2018vitamin; @Dinlen2016association; @Mathyssen2017vitamin; @Esposito2015vitamin; @Jat2017vitamin; @Moriyama2020seasonality].

@@

As is the case with cross-sectional data, it is possible to test the residuals of Model (\ref{eq:sur-sim-block}) for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial]. When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial]. In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor. Spatial analytical approaches were used to understand contagion-difussion processes in the case of infectious disease in general [e.g., @Cliff1998detecting] and the 2003-2004 SARS outbreak in particular [e.g., @Meng2005understanding; @Cao2010spatio]. While we are mindful of the same caveat that the novel SARS-CoV2 may not follow the patterns of previous diseases, there is still evidence from the United States that COVID-19 displays spatial patterns that are consistent with a diffusion process [@Desjardins2020rapid]. For this reason, the spatial lag model is appropriate to model incidence of COVID-19 geographically, since it accounts for potential spatial patterns that result from a process of contagion, as explained next.

### 2020-05-29 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Added some missing references flagged during proof-reading for typesetting.

Whether results from laboratory experiments will hold when the virus
circulates in the community remains uncertain. At a global scale, de
Ángel Solá et al.~(2020) see less risk from SARS-CoV2 in the Caribean
Basin; on the other hand, Coelho et al.~(Coelho et al., 2020) warn that
at least during the exponential phase, expansion of the virus is not
driven by climate. Similarly, whereas Araujo and Naimi (2020) argue that
spread of SARS-CoV2 will likely be constrained by climate, Harbert et
al.~(2020) remain unconvinced that spatial modelling can currently
discriminate the distribution of the disease on the basis of climate, at
least in the United States. Yao et al.~(2020), examined data from China
and came to the conclusion that neither temperature nor ultraviolet
indices had an association with transmission of COVID-19. This is
despite previous research that has linked less exposure to UVB radiation
to higher prevalence and severity of acute respiratory tract infections
(Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
Moriyama, Hugentobler, and Iwasaki 2020).

@@

As is the case with cross-sectional data, it is possible to test the
residuals of Model (\ref{eq:sur-sim-block}) for spatial autocorrelation,
and several tests have been developed to test the null hypothesis of
spatial independence (see López et al., 2014). When the null hypothesis
is rejected, several alternative specifications have been proposed to
include spatial effects (Anselin, 1988, see also 2016). In this paper we
consider a spatial SUR model that incorporates a spatial lag of the
dependent variable as an explanatory factor. Spatial analytical
approaches were used to understand contagion-difussion processes in the
case of infectious disease in general (e.g., Cliff et al., 1998) and the
2003-2004 SARS outbreak in particular (e.g., Meng et al., 2005; Cao et
al., 2010). While we are mindful of the same caveat that the novel
SARS-CoV2 may not follow the patterns of previous diseases, there is
still evidence from the United States that COVID-19 displays spatial
patterns that are consistent with a diffusion process (Desjardins et
al., 2020). For this reason, the spatial lag model is appropriate to
model incidence of COVID-19 geographically, since it accounts for
potential spatial patterns that result from a process of contagion, as
explained next.

@@

\leavevmode\hypertarget{ref-Anselin2016estimation}{}%
Anselin, L., 2016. Estimation and testing in the spatial seemingly
unrelated regression (sur). Geoda Center for Geospatial Analysis;
Computation, Arizona State University. Working Paper 2016-01.

@@

@@

@@

@@

@@

@@

@@

>>>>

Whether results from laboratory experiments will hold when the virus
circulates in the community remains uncertain. At a global scale, de
Ángel Solá et al.~(2020) see less risk from SARS-CoV2 in the Caribean
Basin; on the other hand, Coelho et al.~(Coelho et al., 2020) warn that
at least during the exponential phase, expansion of the virus is not
driven by climate. Similarly, whereas Araujo and Naimi (2020) argue that
spread of SARS-CoV2 will likely be constrained by climate, Harbert et
al.~(2020) remain unconvinced that spatial modelling can currently
discriminate the distribution of the disease on the basis of climate, at
least in the United States. Yao et al.~(2020), examined data from China
and came to the conclusion that neither temperature nor ultraviolet
indices had an association with transmission of COVID-19. This is
despite previous research that has linked less exposure to UVB radiation
to higher prevalence and severity of acute respiratory tract infections
(Dąbrowska-Leonik et al., 2018; Dinlen et al., 2016; Esposito and Lelii,
2015; Jat, 2017; Mathyssen et al., 2017; Moriyama et al., 2020;
Zittermann et al., 2016).

@@

As is the case with cross-sectional data, it is possible to test the
residuals of Model (\ref{eq:sur-sim-block}) for spatial autocorrelation,
and several tests have been developed to test the null hypothesis of
spatial independence (see López et al., 2014). When the null hypothesis
is rejected, several alternative specifications have been proposed to
include spatial effects (Anselin, 1988). In this paper we consider a
spatial SUR model that incorporates a spatial lag of the dependent
variable as an explanatory factor. Spatial analytical approaches were
used to understand contagion-difussion processes in the case of
infectious disease in general (e.g., Cliff et al., 1998) and the
2003-2004 SARS outbreak in particular (e.g., Meng et al., 2005; Cao et
al., 2010). While we are mindful of the same caveat that the novel
SARS-CoV2 may not follow the patterns of previous diseases, there is
still evidence from the United States that COVID-19 displays spatial
patterns that are consistent with a diffusion process (Desjardins et
al., 2020). For this reason, the spatial lag model is appropriate to
model incidence of COVID-19 geographically, since it accounts for
potential spatial patterns that result from a process of contagion, as
explained next.

@@

@@

\leavevmode\hypertarget{ref-Dux105browska2018vitamin}{}%
Dąbrowska-Leonik, N., Bernatowska, E., Pac, M., Filipiuk, W., Mulawka,
J., Pietrucha, B., Heropolitańska-Pliszka, E., Bernat-Sitarz, K.,
Wolska-Kuśnierz, B., Mikołuć, B., 2018. Vitamin d deficiency in children
with recurrent respiratory infections, with or without immunoglobulin
deficiency. Advances in medical sciences 63, 173--178.

@@

\leavevmode\hypertarget{ref-Dinlen2016association}{}%
Dinlen, N., Zenciroglu, A., Beken, S., Dursun, A., Dilli, D., Okumus,
N., 2016. Association of vitamin d deficiency with acute lower
respiratory tract infections in newborns. The Journal of Maternal-Fetal
\& Neonatal Medicine 29, 928--932.

@@

\leavevmode\hypertarget{ref-Esposito2015vitamin}{}%
Esposito, S., Lelii, M., 2015. Vitamin d and respiratory tract
infections in childhood. BMC infectious diseases 15, 487.

@@

\leavevmode\hypertarget{ref-Jat2017vitamin}{}%
Jat, K.R., 2017. Vitamin d deficiency and lower respiratory tract
infections in children: A systematic review and meta-analysis of
observational studies. Tropical doctor 47, 77--84.

@@

\leavevmode\hypertarget{ref-Mathyssen2017vitamin}{}%
Mathyssen, C., Gayan-Ramirez, G., Bouillon, R., Janssens, W., 2017.
Vitamin d supplementation in respiratory diseases: Evidence from
randomized controlled trials. Polish archives of internal medicine 127,
775--784.

@@

\leavevmode\hypertarget{ref-Moriyama2020seasonality}{}%
Moriyama, M., Hugentobler, W.J., Iwasaki, A., 2020. Seasonality of
respiratory viral infections. Annual Review of Virology 7.

@@

\leavevmode\hypertarget{ref-Zittermann2016vitamin}{}%
Zittermann, A., Pilz, S., Hoffmann, H., März, W., 2016. Vitamin d and
airway infections: A european perspective. European journal of medical
research 21, 14.
